Decoupled Fallback: A Portable Single-Pass GPU Scan

### 2025-02-23 paper.tex
Tighten language on barriers inline with Raph's feedback.

\begin{abstract}
  We present \emph{Decoupled Fallback}, a method that enables single-pass \emph{Chained-Scans} to run on hardware without \emph{Forward-Progress Guarantees} (FPG) while avoiding deadlock. Additionally, we introduce a tile state representation for \emph{Chained-Scans} that does not rely on 64-bit atomics or memory fences for correctness, along with a subgroup-size-agnostic intra-workgroup implementation. On an FPG-lacking devices---Apple M1 Max and M3---\emph{Decoupled-Fallback} achieves near-\emph{memcopy} speeds for inclusive prefix sum in WGPU and realizes the full theoretically expected 50\% speedup over the slower \emph{Reduce-then-Scan} approach. We further demonstrate the resilience of \emph{Decoupled-Fallback} against \emph{unfair} schedulers by emulating blocking at rates of up to 50\%, showing that it maintains superior performance over \emph{Reduce-then-Scan} even under extreme contention.
\end{abstract}

@@

\subsubsection{Scan-then-Propagate}
Introduced by Sengupta et al.~\cite{10.5555/1280094.1280110}, \emph{Scan-then-Propagate}~\cite{GPUGems3, Sengupta2011} extends intra-workgroup scans to large inputs by resolving inter-workgroup dependencies in three phases:
\begin{enumerate}
  \item \textbf{Intra-Workgroup (Local) Scan:}  Each workgroup performs an inclusive scan on its assigned work tile, storing local results and reductions in global memory.
  \item \textbf{Spine Scan:} The workgroup reductions, collectively forming the \emph{spine}, are gathered and scanned to compute the \emph{root}, resolving dependencies between work tiles. For large scans, this stage is an insignificant amount of work compared to the first and third.
  \item \textbf{Propagation:} Each workgroup retrieves its portion of the root and adds it to the local scan output.
\end{enumerate}
The absence of inter-workgroup barriers necessitates separate kernel launches for dependency resolution, naturally enforcing this structure. To accommodate arbitrarily large inputs, Sengupta et al. employ recursion, repeatedly applying the three-phase process until the spine fits within a single work tile. Thus, an input of size $n$ and a work tile of size $t$ results in a recursive depth of $\lceil \log_t n \rceil$, and $2\cdot\lceil \log_t n \rceil - 1$ kernel launches. Each recursive step $k$ processes $n/t^k$ tiles, leading to a total of $\sum_{k=1}^{\lceil \log_t n \rceil} n/t^k \approx (n - 1)/(t - 1)$ work tiles over all steps. Since each tile moves $4t$ data ($2t$ local scan, $2t$ propagation), the total global data movement is $O\left(\frac{4t(n - 1)}{t - 1}\right) = O(4n)$.

@@

Merrill and Grimshaw~\cite{Merrill2009} introduced \emph{workgroup raking} to reduce recursion overhead for arbitrarily large inputs. Instead of recursively reducing the spine, the input is partitioned among a fixed number of workgroups, $c$, where $c \geq o$, the total workgroup occupancy. This constrains the spine from a variable, potentially recursive structure to a fixed size $c$, ensuring a recursion depth of at most two. As a result, global memory movement for the spine scan is reduced from $O(n/t)$ to $O(2c)$, and kernel launches are limited to three, yielding a total global memory movement of $O(3n + 4c) = O(3n)$.

Breitbart~\cite{10.5555/2031978.2032029} applied Gupta et al.'s \emph{Persistent-Threads}~\cite{gupta2012} to scan, eliminating kernel launch overheads. Recognizing that raking naturally aligns with this paradigm, Breitbart's method statically \emph{discovers} the workgroup occupancy and launches exactly $o$ kernels. Since multiprocessors are not oversubscribed, workgroups persist for the entire kernel duration, enabling synchronization at an atomics-based inter-workgroup barrier without context switching. While global memory movement remains $O(3n)$, this consolidates the scan into a single kernel, minimizing launch overheads.

\subsubsection{Chained-Scan, Stream-Scan}
First introduced by Yan et al.~\cite{10.1145/2442516.2442539} in \emph{Stream-Scan}, the key innovation of the \emph{Chained-Scan} architecture lies in its hybridization of parallel and serial strategies: achieving parallelism at the intra-workgroup level while minimizing global data movement through serial scan operations at the inter-workgroup level. Instead of using kernel launches as inter-workgroup synchronization points, \emph{Chained-Scans} launch a single kernel that assigns work tiles in a serial order to workgroups, utilizing atomics and bit-packed status flags to ensure coherent views of data. To enforce this serial ordering, work tiles are dynamically assigned using atomic increment operations, rather than arbitrary assignment based on virtualized workgroup indices. This shifts dependency resolution from explicit synchronization via barrier-like structures to implicit synchronization via access order. As a result, each element is read and written exactly once, achieving the theoretical minimum of $O(2n)$.

@@

\subsection{Ideal Scan Architecture}
\emph{Chained Scan} and \emph{Reduce-Then-Scan} represent two contrasting paradigms in scan design. By avoiding explicit pipeline barrier based inter-workgroup synchronization, \emph{Chained Scan} aligns with the oversubscription paradigm, delegating occupancy management to the workgroup scheduler. In contrast, \emph{Reduce-Then-Scan} employs workgroup raking, requiring fixed launches and explicit occupancy control, forcing developers to manually tune launch sizes or implement occupancy discovery techniques.

@@

Given these constraints, oversubscription and \emph{Chained Scan} emerge as the more portable and preferred architecture. Furthermore, its single-kernel design and serial ordering make it suitable for \emph{in-situ} operations, benefiting applications such as data deduplication, sparse data compaction, error correction, and interval merging. However, while \emph{Chained-Scan} with \emph{Decoupled Lookback} appears to be the ideal scan architecture in both theoretical and practical terms, it has a critical limitation: its reliance on a forward-progress guarantee. While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that Apple’s M-series GPUs do not \thomas{I've cut mentioning ARM here due to reclassification, but it kind of reduces the impact. I.e. "Oh it's just Apple that's missing, so this is a vendor hack?"}. Without forward progress guarantees, \emph{Chained-Scan} is not merely inefficient---it risks indefinite deadlock, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.

@@

  \If{$\textit{threadid} = 0$}{
    $\textit{wg\_broadcast} \gets \textnormal{atomicAdd}(\&\textit{tileBump}, 1)$\;
    $\textit{wg\_control} \gets \textnormal{LOCKED}$\;
  }
  \textbf{barrier()}\;

@@

  \textbf{Load}($\textit{in}$, $\textit{tile\_id}$, \textit{local\_vectors})\;
  \textbf{SubgroupRake}($\textit{local\_vectors}$, $\textit{wg\_partials}$)\;
  \textbf{barrier()}\;

  \textbf{WorkgroupWideScan}($\textit{wg\_partials}$)\;
  \textbf{barrier()}\;

  \tcp{Chained Propagation, Decoupled Fallback}
  \textbf{PostLocalReduction}($\textit{spine}$, $\textit{tile\_id}$, $\textit{wg\_partials}$)\;
  \While{$\textit{wg\_control} = \textnormal{LOCKED}$}{
    \textbf{Lookback}($\textit{spine}$, $\textit{tile\_id}$, $\textit{wg\_prev\_reduction}$, $\textit{wg\_control}$)\;
    \textbf{barrier()}\;

    \If{$\textit{wg\_control} = \textnormal{LOCKED}$}{
      \textbf{FallbackReduce}($\textit{in}$, $\textit{wg\_fallback}$)\;
      \textbf{FallbackInsertionAttempt}($\textit{spine}$, $\textit{wg\_fallback}$, $\textit{wg\_prev\_reduction}$, $\textit{wg\_control}$)\;
      \textbf{barrier()}\;
    }
  }

@@

      \textbf{barrier()}\;

@@

\subsection{Tile State Implementation}
In all variants of \emph{Chained-Scan}, there is communication and synchronization between workgroups. The fundamental primitive is \emph{message passing,} meaning that one workgroup posts the result of a communication, then other workgroups read it. This primitive is well studied in the context of memory models~\cite{Alglave0, MCMutants}. \raph{The earliest publication of the MP litmus test appears to be section 2.1 of "Intel 64 Architecture Memory Ordering White Paper". I'm not sure that's the best cite; the MC mutants paper cites https://arxiv.org/pdf/1207.7264} \thomas{Added MC Mutants + the Algave paper. OK? Or should cite more? Alglave has tons of papers.} The key requirement is both control synchronization, meaning the reader does not proceed until the writer sends the message; and data synchronization, meaning that the reader reads the correct value. The latter is especially challenging in relaxed memory models as are employed on GPUs.

Combined control and data synchronization can be accomplished by packing both the tile state and the reduced value into a single word, using relaxed atomic store to send the message, and relaxed atomic load to receive the message, spinning as necessary until the tile state is updated. The semantics of atomics guarantee, even with relaxed memory ordering, that the data is fresh. However, if the tile state and data are stored in two different memory locations, then relaxed atomics are not sufficient and some other sychronization is necessary. On CPUs, it is common to upgrade the memory semantics from relaxed to acquire/release, which then provides the necessary guarantee~\cite[Section 10.2]{intel_sdm_2024} \thomas{I've cited the latest publication by Intel. The section is nearly identical to 2007.} . Unfortunately, on GPUs, acquire/release semantics are neither portable (they are only available on Vulkan, and optional until version 1.3), nor are they necessarily efficient even when they are available; often the implementation involves expensive cache flushes. A device-scoped memory barrier has similar problems. While it would provide the necessary guarantee, it is also not available in WebGPU (or in versions of Metal prior to 3.2), and is expensive.

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width=\linewidth]{graphics/split.pdf}
\end{figure}

In particular, we use two or four 32-bit values, splitting the reduction value into 16-bit segments while reserving the two most significant bits of each 32-bit value to encode the tile state. Although this could also be accomplished with shared memory (and barriers), as an optimization, we use two or four threads from the first subgroup, leveraging subgroup \emph{ballot} to verify state consistency, and subgroup \emph{shuffle} to join and split reductions. With one thread per 16 bits and a minimum subgroup size of four in WebGPU, this method supports both 32-bit and 64-bit values, enabling 64-bit scans without 64-bit atomics or fences.\footnote{Although native 64-bit values are not currently supported by the WebGPU specification, their future inclusion is anticipated.}

@@

\subsection{Trade-offs}
Decoupled Fallback introduces several tradeoffs that must be considered: (1) Without a memory fence, the scan cannot be performed in place and requires a separate output buffer to prevent false-positive fallback operations on post-scanned values, as the tile state may not update before the scan values. (2) For monoids larger than 64 bits, multiple subgroups are required, increasing synchronization overhead due to the need for shared memory and barriers. (3) In certain in-situ compaction algorithms where fallback work-tiles cannot be easily reduced and must be held in registers, Decoupled Fallback may reduce occupancy. (4) Additionally, in compute-bound in-situ compaction algorithms, the cost of fallbacks may outweigh the savings from reduced global memory traffic. (5) Lastly, because the control flow of \emph{Decoupled Fallback} requires barriers, latency incurred during lookback cannot be hidden by workgroup computation.

\subsection{Future Work}
Sorting is one of the most notable beneficiaries of single-pass scan and, by extension, \emph{Decoupled Fallback}. The current state-of-the-art algorithm, Adinet and Merrill’s \emph{Onesweep}~\cite{adinets2022onesweepfastersignificantdigit}, is an 8-bit least-significant-digit radix sort, where single-pass scan minimizes data movement during digit-binning passes. However, due to its compute-intensive nature and large $radixDigit$-sized monoid, it is unclear whether \emph{Decoupled Fallback} maps well to sorting or if new techniques are needed.

>>>>

\begin{abstract}
  We present \emph{Decoupled Fallback}, a method that enables single-pass \emph{Chained-Scans} to run on hardware without \emph{Forward-Progress Guarantees} (FPG) while avoiding deadlock. Additionally, we introduce a tile state representation for \emph{Chained-Scans} that does not rely on 64-bit atomics or memory barriers for correctness, along with a subgroup-size-agnostic intra-workgroup implementation. On an FPG-lacking devices---Apple M1 Max and M3---\emph{Decoupled-Fallback} achieves near-\emph{memcopy} speeds for inclusive prefix sum in WGPU and realizes the full theoretically expected 50\% speedup over the slower \emph{Reduce-then-Scan} approach. We further demonstrate the resilience of \emph{Decoupled-Fallback} against \emph{unfair} schedulers by simulating blocking at rates of up to 50\%, showing that it maintains superior performance over \emph{Reduce-then-Scan} even under extreme contention.
\end{abstract}

@@

\subsubsection{Scan-then-Propagate}
Introduced by Sengupta et al.~\cite{10.5555/1280094.1280110}, \emph{Scan-then-Propagate}~\cite{GPUGems3, Sengupta2011} extends intra-workgroup scans to large inputs by resolving inter-workgroup dependencies in three phases:
\begin{enumerate}
  \item \textbf{Intra-Workgroup (Local) Scan:}  Each workgroup performs an inclusive scan on its assigned work tile, storing local results and reductions in global memory.
  \item \textbf{Spine Scan:} The workgroup reductions, collectively forming the \emph{spine}, are gathered and scanned to compute the \emph{root}, resolving dependencies between work tiles. For large scans, this stage is an insignificant amount of work compared to the first and third.
  \item \textbf{Propagation:} Each workgroup retrieves its portion of the root and adds it to the local scan output.
\end{enumerate}
The absence of inter-workgroup barriers necessitates separate kernel launches and pipeline barriers for dependency resolution, naturally enforcing this structure. To accommodate arbitrarily large inputs, Sengupta et al. employ recursion, repeatedly applying the three-phase process until the spine fits within a single work tile. Thus, an input of size $n$ and a work tile of size $t$ results in a recursive depth of $\lceil \log_t n \rceil$, and $2\cdot\lceil \log_t n \rceil - 1$ kernel launches. Each recursive step $k$ processes $n/t^k$ tiles, leading to a total of $\sum_{k=1}^{\lceil \log_t n \rceil} n/t^k \approx (n - 1)/(t - 1)$ work tiles over all steps. Since each tile moves $4t$ data ($2t$ local scan, $2t$ propagation), the total global data movement is $O\left(\frac{4t(n - 1)}{t - 1}\right) = O(4n)$.

@@

Merrill and Grimshaw~\cite{Merrill2009} introduced \emph{workgroup raking} to reduce recursion overhead for arbitrarily large inputs. Instead of recursively reducing the spine, the input is partitioned among a fixed number of workgroups, $c$, where $c \geq o$, the total workgroup occupancy. This constrains the spine from a variable, potentially recursive structure to a fixed size $c$, ensuring a recursion depth of at most two. As a result, global memory movement for the spine scan is reduced from $O(n/t)$ to $O(2c)$, and kernel launches and pipeline barriers are limited to three, yielding a total global memory movement of $O(3n + 4c) = O(3n)$.

Breitbart~\cite{10.5555/2031978.2032029} applied Gupta et al.'s \emph{Persistent-Threads}~\cite{gupta2012} to scan, eliminating kernel launch overheads. Recognizing that raking naturally aligns with this paradigm, Breitbart's method statically \emph{discovers} the workgroup occupancy and launches exactly $o$ kernels. Since multiprocessors are not oversubscribed, workgroups persist for the entire kernel duration, enabling synchronization at an atomics-based inter-workgroup memory and control barrier without kernel launches or pipeline barriers. While global memory movement remains $O(3n)$, this consolidates the scan into a single kernel, minimizing launch and pipeline overheads.

\subsubsection{Chained-Scan, Stream-Scan}
First introduced by Yan et al.~\cite{10.1145/2442516.2442539} in \emph{Stream-Scan}, the key innovation of the \emph{Chained-Scan} architecture lies in its hybridization of parallel and serial strategies: achieving parallelism at the intra-workgroup level while minimizing global data movement through serial scan operations at the inter-workgroup level. Instead of using kernel launches and pipeline barriers as inter-workgroup synchronization points, \emph{Chained-Scans} launch a single kernel that assigns work tiles in a serial order to workgroups, utilizing atomics and bit-packed status flags to ensure coherent views of data. To enforce this serial ordering, work tiles are dynamically assigned using atomic increment operations, rather than arbitrary assignment based on virtualized workgroup indices. This shifts dependency resolution from explicit synchronization via barriers to implicit synchronization via access order. As a result, each element is read and written exactly once, achieving the theoretical minimum of $O(2n)$.

@@

\subsection{Ideal Scan Architecture}
\emph{Chained Scan} and \emph{Reduce-Then-Scan} represent two contrasting paradigms in scan design. By eliding the barrier-idiom, \emph{Chained Scan} aligns with the oversubscription paradigm, delegating occupancy management to the workgroup scheduler. In contrast, \emph{Reduce-Then-Scan} employs workgroup raking, requiring fixed launches and explicit occupancy control, forcing developers to manually tune launch sizes or implement occupancy discovery techniques.

@@

Given these constraints, oversubscription and \emph{Chained Scan} emerge as the more portable and preferred architecture. Furthermore, its single-kernel design and serial ordering make it suitable for \emph{in-situ} operations, benefiting applications such as data deduplication, sparse data compaction, error correction, and interval merging. However, while \emph{Chained-Scan} with \emph{Decoupled Lookback} appears to be the ideal scan architecture in both theoretical and practical terms, it has a critical limitation: its reliance on a forward-progress guarantee. While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that ARM and Apple’s M-series GPUs do not. Without forward progress guarantees, \emph{Chained-Scan} is not merely inefficient---it risks indefinite deadlock, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.

@@

  \If{$\textit{threadid} = 0$}{
    $\textit{wg\_broadcast} \gets \textnormal{atomicAdd}(\&\textit{tileBump}, 1)$\;
    $\textit{wg\_control} \gets \textnormal{LOCKED}$\;
  }
  \textbf{workgroupBarrier()}\;

@@

  \textbf{Load}($\textit{in}$, $\textit{tile\_id}$, \textit{local\_vectors})\;
  \textbf{SubgroupRake}($\textit{local\_vectors}$, $\textit{wg\_partials}$)\;
  \textbf{workgroupBarrier()}\;

  \textbf{WorkgroupWideScan}($\textit{wg\_partials}$)\;
  \textbf{workgroupBarrier()}\;

  \tcp{Chained Propagation, Decoupled Fallback}
  \textbf{PostLocalReduction}($\textit{spine}$, $\textit{tile\_id}$, $\textit{wg\_partials}$)\;
  \While{$\textit{wg\_control} = \textnormal{LOCKED}$}{
    \textbf{Lookback}($\textit{spine}$, $\textit{tile\_id}$, $\textit{wg\_prev\_reduction}$, $\textit{wg\_control}$)\;
    \textbf{workgroupBarrier()}\;

    \If{$\textit{wg\_control} = \textnormal{LOCKED}$}{
      \textbf{FallbackReduce}($\textit{in}$, $\textit{wg\_fallback}$)\;
      \textbf{FallbackInsertionAttempt}($\textit{spine}$, $\textit{wg\_fallback}$, $\textit{wg\_prev\_reduction}$, $\textit{wg\_control}$)\;
      \textbf{workgroupBarrier()}\;
    }
  }

@@

      \textbf{workgroupBarrier()}\;

@@

\subsection{Tile State Implementation}
In all variants of \emph{Chained-Scan}, there is communication and synchronization between workgroups. The fundamental primitive is \emph{message passing,} meaning that one workgroup posts the result of a communication, then other workgroups read it. This primitive is well studied in the context of memory models~\cite{Alglave0, MCMutants}. The key requirement is both control synchronization, meaning the reader does not proceed until the writer sends the message; and data synchronization, meaning that the reader reads the correct value. The latter is especially challenging in relaxed memory models as are employed on GPUs.

Combined control and data synchronization can be accomplished by packing both the tile state and the reduced value into a single word, using relaxed atomic store to send the message, and relaxed atomic load to receive the message, spinning as necessary until the tile state is updated. The semantics of atomics guarantee, even with relaxed memory ordering, that the data is fresh. However, if the tile state and data are stored in separate memory locations, relaxed atomics are insufficient because the compiler is free to reorder stores, so one update does not imply the other. On CPUs, it is common to upgrade the memory semantics from relaxed to acquire/release, which then provides the necessary guarantee~\cite[Section 10.2]{intel_sdm_2024}. Unfortunately, on GPUs, acquire/release semantics are neither portable (they are only available on Vulkan, and optional until version 1.3), nor are they necessarily efficient even when they are available; often the implementation involves expensive cache flushes. A device-scoped memory barrier has similar problems. While it would provide the necessary guarantee, it is also not available in WebGPU (or in versions of Metal prior to 3.2), and is expensive.

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width=\linewidth]{graphics/split.pdf}
  \caption{Splitting one 32-bit value into two 32-bit values with bit-packed tile state.}
\end{figure}

In particular, we use two or four 32-bit values, splitting the reduction value into 16-bit segments while reserving the two most significant bits of each 32-bit value to encode the tile state. Although this could also be accomplished with shared memory (and workgroup barriers), as an optimization, we use two or four threads from the first subgroup, leveraging subgroup \emph{ballot} to verify state consistency, and subgroup \emph{shuffle} to join and split reductions. With one thread per 16 bits and a minimum subgroup size of four in WebGPU, this method supports both 32-bit and 64-bit values, enabling 64-bit scans without 64-bit atomics or memory barriers.\footnote{Although native 64-bit values are not currently supported by the WebGPU specification, their future inclusion is anticipated.}

@@

\subsection{Trade-offs}
\emph{Decoupled Fallback} introduces several tradeoffs that must be considered: (1) Without a memory barrier, the scan cannot be performed in place and requires a separate output buffer to prevent false-positive fallback operations on post-scanned values, as the tile state may not update before the scan values. (2) For monoids larger than 64 bits, multiple subgroups are required, increasing synchronization overhead due to the need for shared memory and barriers. (3) In certain in-situ compaction algorithms where fallback work-tiles cannot be easily reduced and must be held in registers, Decoupled Fallback may reduce occupancy. (4) Additionally, in compute-bound in-situ compaction algorithms, the cost of fallbacks may outweigh the savings from reduced global memory traffic. (5) Lastly, because the control flow of \emph{Decoupled Fallback} requires barriers, latency incurred during lookback cannot be hidden by workgroup computation.

\subsection{Future Work}
Sorting is one of the most notable beneficiaries of single-pass scan and, by extension, \emph{Decoupled Fallback}. The current state-of-the-art algorithm, Adinets and Merrill’s \emph{Onesweep}~\cite{adinets2022onesweepfastersignificantdigit}, is an 8-bit least-significant-digit radix sort, where single-pass scan minimizes data movement during digit-binning passes. However, due to its compute-intensive nature and large $radixDigit$-sized monoid, it is unclear whether \emph{Decoupled Fallback} will map well to sorting or if new techniques are needed.
